Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary

### 2024-02-06 manu-draft.qmd
working on BF revisions

---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary"
bibliography: refs.bib
csl: estuaries-and-coasts.csl
author:
  - Marcus W Beck:
      email: mbeck@tbep.org
      orcid: 0000-0002-4996-0059
      institute: [TBEP]
      correspondence: true
  - Kerry Flaherty-Walia:
      email: kfwalia@tbep.org
      orcid: 0000-0002-6032-0570
      institute: [TBEP]
      correspondence: false
  - Sheila Scolaro:
      email: sscolaro@tbep.org
      orcid: 0000-0003-3220-3737
      institute: [TBEP]
      correspondence: false
  - Maya C Burke:
      email: mburke@tbep.org
      orcid: 0000-0002-3847-6501      
      institute: [TBEP]
      correspondence: false
  - Bradley T Furman:
      email: Bradley.Furman@myfwc.com
      orcid: 0000-0002-9284-8388
      institute: [FWRI]
      correspondence: false
  - David J Karlen: 
      email: karlen@epchc.org
      orcid: 0000-0002-1432-9099
      institute: [EPCHC]
      correspondence: false
  - Chris Pratt: 
      email: prattc@epchc.org
      institute: [EPCHC]
      correspondence: false
  - Edward T Sherwood:
      email: esherwood@tbep.org
      orcid: 0000-0001-5330-302X
      institute: [TBEP]
      correspondence: false

institute:
  - TBEP: Tampa Bay Estuary Program, 263 13th Ave S., Suite 350, St. Petersburg, Florida 33701 USA
  - FWRI: Florida Fish and Wildlife Conservation Commission, Florida Fish and Wildlife Research Institute, 100 8th Avenue SE, St. Petersburg, FL 33701, USA
  - EPCHC: Environmental Protection Commission of Hillsborough County, 3629 Queen Palm Drive, Tampa, FL 33619, USA

@@

Climate change is an existential threat to natural resources that sustain human health and the environment. In Tampa Bay, Florida (USA), seagrass coverage has declined since 2016 by nearly a third (11,518 acres), despite sustained reductions of nitrogen loads supportive of light environments for growth. Changing physical water quality conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be linked to climate stress. Three ambient water quality datasets of varying sampling design and coverage were evaluated to characterize physicochemical environments in Tampa Bay and the potential relationships with seagrass change. Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 PSU per year. These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay during the summer months. Simple linear models provided limited evidence that recent seagrass declines are associated with hotter and fresher conditions. Trends in warming and increased precipitation in the region are likely to continue, further creating suboptimal conditions for seagrasses in Tampa Bay. These results should compel resource managers to consider the likelihood that reduced resilience of estuarine resources due to shifting ecological baselines driven by additional climate change drivers will complicate long-standing management paradigms. While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be confounded by other cliamte change drivers and warrant additional, complementary interventions to maintain ecosystem health into the future.

@@

The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale. Seagrasses are fundamental indicators of coastal ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services [@fourqurean2012; @orth2020; @orth2023]. Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021]. These losses are comparable to, if not more significant than, other interconnected critical coastal environments such as mangroves and salt marshes [@duarte2008a], which threatens the natural and economic services that these habitats provide. Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @hall1999; @orth2006; @burkholder2007]. Furthermore, natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020]. Contemporary management actions and interventions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].

The sustained coverage or restoration of seagrasses in stressed coastal environments requires management actions that support vegetative growth, reproduction, and coverage expansion. A long-standing approach adopted by numerous management entities has been the control of external nutrient inputs in systems where excessive algal growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @han2014; @greening2014]. There are limited examples of successful recovery of seagrass through control of nutrient inputs alone, primarily because of the difficulty in identifying and regulating both point and diffuse non-point sources. Notable exceptions include Tampa Bay on the west coast of Florida [@greening2014] and the much larger Chesapeake Bay on the east Atlantic US coast [@lefcheck2018], where both showed significant increase in seagrass areal coverage through sustained and long-term reductions in external nutrient loads. Cooperation among management, regulatory, public, and private sectors were critical aspects of both examples [@tango2016; @sherwood2016]. In other cases, the reversal of seagrass losses through nutrient reductions alone may not be possible because of system hysteresis, where the path to recovery is not the same as the path to decline [@maxwell2016]. Complementary management actions, in addition to nutrient reductions, are needed in these situations.

Climate change has complicated the understanding of ecosystem response to conventional stressors, presenting new challenges and expectations for how ecological resources will respond to management actions [@statham2012a; @sherwood2013]. In addition to sea-level rise, the most anticipated effects of climate change in coastal environments are increasing temperature and alteration of precipitation patterns. These changes will profoundly alter physicochemical habitats, creating suboptimal or uninhabitable conditions for many species [@madeira2012; @lefcheck2017; @hensel2023]. @lefcheck2017 evaluated interactive effects of water clarity and rising temperatures on seagrasses with over 30 years of data in Chesapeake Bay. The effects of decreased water clarity for seagrasses in shallow habitats were exacerbated by acute warming related to climate change, demonstrating a concerning synergy of multiple stressors. @moreno-marin2018 produced similar results using a multifactorial experiment that considered temperature, light, and nutrient (nitrogen) availability for North Sea eelgrass (*Zostera marina*). Species shifts are also expected to occur as changing physical conditions decrease the competitive advantages of historically abundant species. Again in Chesapeake Bay, the abundance of widgeongrass (*Ruppia maritima*) has been mediated by nutrient reductions, while also outcompeting and replacing formerly dominant *Z. marina* given higher tolerance to increased temperatures [@hensel2023]. Changing frequency and severity of precipitation patterns may further alter the distribution and abundance of seagrasses by pushing species outside of optimal salinity tolerance ranges [@rasheed2011; @webster2021]. These changes may produce hypersaline conditions under periods of prolonged drought or more freshwater conditions with increased storm events. Seagrasses may respond differently under individual or multiple stressors [@lefcheck2017; @zhang2023], suggesting place-based assessments of response to climate change provide valuable insights for managers responsible for maintaining ecological resources.

In Tampa Bay, Florida, seagrasses are a primary indicator of bay health that have been the focus of management efforts over the last thirty years [@sherwood2017]. Through successful reduction of external nitrogen loads, seagrasses have recovered from a low, system-wide coverage in the 1980s to an all-time high in 2016 of 41,655 acres [16,857 ha, @greening2006; @greening2014]. Seagrass distribution and abundance has been dynamic throughout this period, overall responding positively to increases in water clarity with nutrient load reductions, while also demonstrating more short-term variability in response to regional climate events [@greening2006]. Since 2016, seagrass areal cover has decreased by nearly 1/3 despite relatively stable water quality [@tbep0696; @Beck20a]. Factors that have influenced this recent decline are unknown and the effects of climate change drivers on physical water quality conditions have been implicated as potential stressors. Following global trends, recent work has demonstrated a broad long-term trend of increasing water temperature in Tampa Bay, although at a relatively coarse scale [@nickerson2023]. Changing salinity has not been well-described, nor have potential links of changing salinity and temperature with recent seagrass change. Tampa Bay is rich with historical data that can be used to evaluate long-term trends. This information can fill a critical knowledge gap that can inform regional management activities, while also demonstrating the confounding effects of climate change with ecosystem response to conventional stressors.

@@

Tampa Bay is divided into distinct sub-segments defined by physical and natural boundaries to assist with water quality management activities [@lewis1985]: Old Tampa Bay (OTB) in the northwest; Hillsborough Bay (HB) in the northeast; Middle Tampa Bay (MTB); and Lower Tampa Bay (LTB) that connects to the Gulf of Mexico (Figure [-@fig-map]a). Old Tampa Bay and Hillsborough Bay have historically had the most degraded water quality primarily from direct external nutrient inputs from wastewater and stormwater [@greening2014]. Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small tributaries and manmade flood control conveyances [@tbep0623]. Notably, Old Tampa Bay has restricted circulation from multiple land bridges associated with causeways that traverse the bay [@sherwood2015; @luther2022]. Recurring seasonal harmful algal blooms of the dinoflagellete *Pyrodinium bahamense* have contributed to exceedances of the chlorophyll-a regulatory standard in Old Tampa Bay [@lopez2023]. By comparison, water quality conditions in Middle Tampa Bay and Lower Tampa Bay are generally better than the upper two bay segments primarily from more frequent water exchanges with the Gulf of Mexico and lower nutrient loading [@tbep0623]. All bay segments are relatively shallow, with a baywide mean depth of approximately 3 m. Light penetration typically reaches bottom habitats under current conditions, although seagrasses were historically limited by high phytoplankton production that affected light environments[@greening2014; @johansson15].

@@

The second goal of the analysis was to evaluate if seagrass changes were linked to long-term changes in water temperature and salinity. The conceptual model for evaluating these changes describes the niche space where seagrass growth and reproduction is hypothesized to be greatest within optimal ranges for forcing factors that are present in the environment [@hutchinson1957; @vandermeer1972]. In the simplest form, this can conceptualized as a bell curve with optimal conditions defined within a range of values for a single parameter (e.g., minimum and maximum temperatures where a species is typically observed), where reduced growth or mortality is observed outside of these ranges. Because both water temperature and salinity were evaluated, the same model can be conceptualized in two-dimensional space ([@fig-concept]). Seagrass growth can be limited when temperature is below or above the optimum range, when salinity is below or above the optimum range, or when both temperature and salinity conditions are outside of the optimum range. Based on the results of the trend tests, we hypothesized that seagrasses are likely stressed by both high temperature and low salinity (bottom right, @fig-concept). Although the optimal niche space can be defined in multiple dimensions for many parameters, we focus on water temperature and salinity given that other dominant forcing factors, i.e., light availability, have generally not been limiting for growth in recent years.

A fundamental challenge describing niche space is identifying the boundaries for optimal conditions. In Tampa Bay, three dominant seagrass species occur: *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinum* (turtle grass) [@lewis1985]. These species co-occur often in mixed beds throughout the bay, although some differences in abundance are observed across salinity ranges. Shoal grass is tolerant of a wide range of salinity [@lirman2003], but is more abundant in oligo/mesohaline portions of Tampa Bay. Conversely, turtle grass is less tolerant of low salinity and is more abundant in more euryhaline conditions near the mouth of Tampa Bay. Reported salinity ranges for each of these species varies depending on location, season, and other co-occurring factors like temperature [@phillips1960; @mcmillan1967; @zieman1975; @lewis1985], although most studies place lower limits of salinity in the range of 15-25 ppt. Optimal temperature ranges are similar between these temperate-tropical species, with reduced growth observed at temperatures above 30$^\circ$ C [@zieman1975; @lewis1985].

Because of the uncertainty in defining *in situ* thresholds for optimal temperature and salinity ranges, multiple thresholds were evaluated to describe the potential for stress and how it may be related to changes in seagrass. Distinctions were not made between species, primarily due to lack of consensus between studies and likely site-specific ranges that affect seagrass growth in Tampa Bay. First, we developed metrics of potential temperature and salinity stress by quantifying the maximum number of continuous days each year when temperature was above or salinity was below a given threshold. This approach assumed that stress could be observed based on duration of exposure (i.e., maximum number of continuous days each year) relative to a threshold that may or may not be outside of the optimum range for seagrasses. These metrics were quantified from the monthly long-term observations in the EPC data. To quantify daily counts each year, a continuous prediction of temperature and salinity over time at each of 45 stations was estimated using Generalized Additive Models (GAMs) fit to temperature or salinity with a single predictor for decimal year [@wood17]. Model fit for each station was considered sufficient to calculate daily predictions to assess potential stressor metrics (Figure S5, R$^2$ ranged from `r fitest$temp$minr2` to `r fitest$temp$maxr2` for temperature models, `r fitest$sal$minr2` to `r fitest$sal$maxr2` for salinity models, Tables S1, S2).

@@

For comparison to seagrass, the annual metrics calculated from the EPC data were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends above. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment from 1998 to 2022. The stressor metrics were compared to frequency occurrence (all species) each year by bay segment. Generalized Linear regression Models (GLMs) were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre-2016, decline post-2016). Two models were evaluated, one with the temperature and salinity metrics together and another with the both metric. Interactions were included between temperature, salinity, and time period or the both metric and time period, depending on the model. Bay segment was included as a main effect without interactions. The most parsimonious models were identified by forward and backward selection of the independent variables and comparison with AIC values [@sakamoto1986; @venables2002b]. All models excluded Lower Tampa Bay because of minimal seagrass change over time.

@@

The EPC dataset was also used to provide detailed information on station-level trends from 1975 to present (@fig-kendall, see Figure S4 for 1998 to 2022). All stations had significantly increasing temperature and decreasing salinity from 1975 to present, excluding one station in HB that did not have significantly decreasing bottom salinity (Figure [-@fig-kendall]a). Seasonally, most stations had more significant trends in the summer, early fall period for both temperature and salinity (Figure [-@fig-kendall]b, c), although some variation was observed between bay segments. Temperature trends were more often observed in the summer or early fall for the upper bay segments (OTB, HB), whereas the lower bay segments (MTB, LTB) had more seasonally homogeneous trends. Seasonal trends in salinity did not vary as much between segments, although OTB and LTB had more stations with significantly decreasing trends than the other bay segments.

@@

Linear mixed-effects models for the EPC data showed similar trends in each bay segment for the number of days when temperature was above different thresholds. All of the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed significantly increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$C (Table S3). The estimated slopes for the number of days when temperature was above 30 $^\circ$C varied from 0.9 days per year in LTB to 1.1 days per year in MTB. Likewise, the average number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied from 8 (OTB, HB) to 21 (LTB) days in 1975 to 55 (HB) to 63 (LTB) days in 2022 (Table S5, see Table S6 for 1998 to 2022). Trends for the other two temperature thresholds were similar.

The salinity models were less similar between bay segments compared to the temperature models, primarily because of the natural variation in salinity along the bay's longitudinal axis (Tables S3, S4). None of the bay segments had significantly increasing number of days per year when salinity was below 15 ppt. The only significant model for MTB was the number of days when salinity was below 25 ppt, which was increasing at a rate of 0.7 days per year. Both HB and OTB had significantly increasing number of days per year when salinity was below 20 and 25 ppt, although the slopes varied such that the rates in OTB were nearly double those in HB. The number of days per year when salinity was below 25 ppt in OTB increased by 2 days per year, whereas the number of days in HB increased by 0.9 days per year from 1975 to present. Likewise, the average number of days per year when salinity was continuously below 25 ppt at the beginning and end of the period of record was 128 (OTB) and 131 (HB) days in 1975 and 171 (HB) and 219 (OTB) days in 2022 (Table S5, see Table S6 for 1998 to 2022).

@@

Global increases in temperature and precipitation related to climate change have contributed to widespread alteration of the structure and functioning of natural environments [@osland2015; @oliver2018]. At a local scale, these changes have been demonstrated herein for Tampa Bay where the evaluation of long-term trends in water temperature and salinity showed clear changes mirrored by long-term changes in air temperature and precipitation. Over the previous fifty years, Tampa Bay has consistently trended towards hotter and fresher conditions, where water temperature has approximately increased by 0.03 - 0.07 $^\circ$C per year and salinity has decreased by 0.04 - 0.2 ppt per year. These changes were demonstrated in three long-term datasets with different sampling methods and periods of record. Understandably, the trends were most consistently observed in the dataset with the longest period of record (EPC) covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season at locations farther from the mouth of the Bay. These long-term changes manifested into consistent trends in select stressor metrics; the continuous number of days each year when temperature was above and/or salinity was below defined thresholds consistently increased across the period of record.

>>>>

---
format: 
  docx:
    reference-doc: my_styles.docx
    number-sections: true
title: "Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary"
bibliography: refs.bib
csl: estuaries-and-coasts.csl
author:
  - Marcus W Beck:
      email: mbeck@tbep.org
      orcid: 0000-0002-4996-0059
      institute: [TBEP]
      correspondence: true
  - Kerry Flaherty-Walia:
      email: kfwalia@tbep.org
      orcid: 0000-0002-6032-0570
      institute: [TBEP]
      correspondence: false
  - Sheila Scolaro:
      email: sscolaro@tbep.org
      orcid: 0000-0003-3220-3737
      institute: [TBEP]
      correspondence: false
  - Maya C Burke:
      email: mburke@tbep.org
      orcid: 0000-0002-3847-6501      
      institute: [TBEP]
      correspondence: false
  - Bradley T Furman:
      email: Bradley.Furman@myfwc.com
      orcid: 0000-0002-9284-8388
      institute: [FWRI]
      correspondence: false
  - David J Karlen: 
      email: karlen@epchc.org
      orcid: 0000-0002-1432-9099
      institute: [EPCHC]
      correspondence: false
  - Chris Pratt: 
      email: prattc@epchc.org
      institute: [EPCHC]
      correspondence: false
  - Chris Anastasiou: 
      email: chris.anastasiou@swfwmd.state.fl.us
      institute: [SWFWMD]
      correspondence: false      
  - Edward T Sherwood:
      email: esherwood@tbep.org
      orcid: 0000-0001-5330-302X
      institute: [TBEP]
      correspondence: false

institute:
  - TBEP: Tampa Bay Estuary Program, 263 13th Ave S., Suite 350, St. Petersburg, Florida 33701 USA
  - FWRI: Florida Fish and Wildlife Conservation Commission, Florida Fish and Wildlife Research Institute, 100 8th Avenue SE, St. Petersburg, FL 33701, USA
  - EPCHC: Environmental Protection Commission of Hillsborough County, 3629 Queen Palm Drive, Tampa, FL 33619, USA
  - SWFWMD: Southwest Florida Water Management District, 7601 US Highway 301 North, Tampa, FL 33637, USA

@@

Seagrasses have long been a focal point for management efforts aimed at restoring ecosystem health in estuaries worldwide. In Tampa Bay, Florida (USA), seagrass coverage has declined since 2016 by nearly a third (11,518 acres), despite sustained reductions of nitrogen loads supportive of light environments for growth. Changing physical water quality conditions related to climate change may be stressing seagrasses beyond their optimal growth ranges, requiring an assessment to determine if this decline can be linked to climate stress. Three ambient water quality datasets of varying sampling design and coverage were evaluated to characterize physicochemical environments in Tampa Bay and the potential relationships with seagrass change. Over the last fifty years, Tampa Bay has become hotter and fresher with water temperature increasing by 0.03 - 0.07 $^\circ$C per year and salinity decreasing by 0.04 - 0.2 ppt per year. These changes varied spatially and seasonally, with the most dramatic changes observed in the upper bay during the summer months. Simple linear models provided a weight-of-evidence that recent seagrass declines are somewhat associated with hotter and fresher conditions. Trends in warming and increased precipitation in the region are likely to continue, further creating suboptimal conditions for seagrasses in Tampa Bay. These results should compel resource managers to consider the likelihood that reduced resilience of estuarine resources due to shifting ecological baselines driven by additional climate change drivers will complicate long-standing management paradigms. While conventional management approaches that focus on limiting nutrient loads should be continued, their future effectiveness may be confounded by other cliamte change drivers and warrant additional, complementary interventions to maintain ecosystem health into the future.

@@

The monitoring and management of seagrasses in coastal environments has received substantial attention at a global scale. Seagrasses are fundamental indicators of coastal ecosystem health [@roca2016; @orth2017], while also serving as foundational species that provide numerous ecosystem services [@fourqurean2012; @orth2020; @orth2023]. Seagrasses have been in global decline with rapid development of coastal environments, particularly in the latter half of the 20th century with accelerating losses estimated at a rate of 110 km$^2$ yr$^{-1}$ since the 1980s [@waycott2009; @dunic2021]. These losses are comparable to, if not more significant than, other interconnected critical coastal environments such as mangroves and salt marshes [@duarte2008a], which threatens the natural and economic services that these habitats provide. Losses have been attributed to numerous stressors including decline in light environments with nutrient enrichment, sedimentation, and physical disturbance primarily from human activities [@duarte1995; @hall1999; @orth2006; @burkholder2007]. Furthermore, natural disturbances such as storm events and disease have also been implicated [@robblee1991; @tomasko2020]. Contemporary management actions aimed at mitigating loss and ultimately supporting restoration require adaptive approaches to address the effects of multiple stressors that have contributed to seagrass decline [@dunic2023].

The sustained coverage or restoration of seagrasses in coastal environments requires environmental conditions that support vegetative growth, reproduction, and coverage expansion. A long-standing approach adopted by numerous management entities has been the control of external nutrient inputs in systems where excessive algal growth has created poor light environments for seagrasses [@boesch2001; @greening2006; @han2014; @greening2014]. There are limited examples of successful recovery of seagrass through control of nutrient inputs alone, primarily because of the difficulty in identifying and regulating both point and diffuse non-point sources. Notable exceptions include Tampa Bay on the west coast of Florida [@greening2014] and the much larger Chesapeake Bay on the east Atlantic US coast [@lefcheck2018], where both showed significant increase in seagrass areal coverage through sustained and long-term reductions in external nutrient loads. Cooperation among management, regulatory, public, and private sectors were critical aspects of both examples [@tango2016; @sherwood2016]. In other cases, the reversal of seagrass losses through nutrient reductions alone may not be possible because of system hysteresis, where the path to recovery is not the same as the path to decline [@maxwell2016]. Complementary management actions, in addition to nutrient reductions, are needed in these situations.

Climate change has complicated the understanding of ecosystem response to conventional stressors, presenting new challenges and expectations for how ecological resources will respond to management actions [@statham2012a; @sherwood2013]. In addition to sea-level rise, the most anticipated effects of climate change in coastal environments are increased temperature and altered precipitation patterns. These changes will profoundly alter physicochemical habitats, creating suboptimal or uninhabitable conditions for many species [@madeira2012; @lefcheck2017; @hensel2023]. @lefcheck2017 evaluated interactive effects of water clarity and rising temperatures on seagrasses with over 30 years of data in Chesapeake Bay. The effects of decreased water clarity for seagrasses in shallow habitats were exacerbated by acute warming related to climate change, demonstrating a concerning synergy of stressors most likely to affect Tampa Bay. @moreno-marin2018 produced similar results using a multifactorial experiment that considered temperature, light, and nutrient (nitrogen) availability for North Sea eelgrass (*Zostera marina*). Species shifts are also expected to occur as changing physical conditions decrease the competitive advantages of historically abundant species. In Chesapeake Bay, the abundance of widgeongrass (*Ruppia maritima*) has responded positively to nutrient reduction and is replacing the formerly dominant but now heat-stressed [*Z. marina*, @hensel2023]. Changing frequency and severity of precipitation patterns may further alter salinity regimes and with it the distribution and abundance of seagrasses throughout the estuary [@rasheed2011; @webster2021]. These changes may produce hypersaline conditions under periods of prolonged drought or more freshwater conditions with increased storm events. Seagrass species may respond differently under individual or multiple stressors and each watershed will respond differently to climate change [@lefcheck2017; @zhang2023], suggesting place-based empirical assessments will be needed to properly inform management decisions.

In Tampa Bay, Florida, seagrasses are a primary indicator of bay health and have been the focus of management efforts for the last thirty years [@sherwood2017]. Through successful reduction of external nitrogen loads, seagrasses have recovered from a low, system-wide coverage in the 1980s to an all-time high in 2016 of 41,655 acres [16,857 ha, @greening2006; @greening2014]. Seagrass distribution and abundance has been dynamic throughout this period, overall responding positively to increases in water clarity with nutrient load reductions, while also demonstrating more short-term variability in response to regional climate events [@greening2006]. Since 2016, seagrass areal cover has decreased by nearly 1/3 despite relatively stable water quality [@tbep0696; @Beck20a]. Factors that have influenced this recent decline are unknown and the effects of climate change drivers on physical water quality conditions have been implicated as potential stressors. Following global trends, recent work has demonstrated a broad long-term trend of increasing water temperature in Tampa Bay, although at a relatively coarse scale [@nickerson2023]. Changing salinity has not been well-described, nor have potential links of changing salinity and temperature with recent seagrass change. Tampa Bay is rich with historical data that can be used to evaluate long-term trends. This information can fill a critical knowledge gap that can inform regional management activities, while also demonstrating the confounding effects of climate change with ecosystem response to conventional stressors.

@@

Tampa Bay is divided into distinct sub-segments defined by physical and natural boundaries to assist with water quality management activities [@lewis1985]: Old Tampa Bay (OTB) in the northwest; Hillsborough Bay (HB) in the northeast; Middle Tampa Bay (MTB); and Lower Tampa Bay (LTB) that connects to the Gulf of Mexico (Figure [-@fig-map]a). Old Tampa Bay and Hillsborough Bay have historically had the most degraded water quality primarily from direct external nutrient inputs from wastewater and stormwater [@greening2014]. Hydrologic conditions vary between the two, such that Hillsborough Bay receives a majority of direct surface water inflow from the Hillsborough and Alafia rivers, whereas Old Tampa Bay receives much less inflow with a majority from multiple small channelized tributaries and manmade flood control conveyances [@tbep0623]. Notably, Old Tampa Bay has restricted circulation from multiple land bridges associated with causeways that traverse the bay [@sherwood2015; @luther2022]. Recurring seasonal harmful algal blooms of the dinoflagellete *Pyrodinium bahamense* have contributed to exceedances of the chlorophyll-a regulatory standard in Old Tampa Bay [@lopez2023]. By comparison, water quality conditions in Middle Tampa Bay and Lower Tampa Bay are generally better than the upper two bay segments primarily from more frequent water exchanges with the Gulf of Mexico and lower nutrient loading [@tbep0623]. All bay segments are relatively shallow, with a baywide mean depth of approximately 3 m. Light penetration typically reaches bottom habitats under current conditions, although seagrasses were historically limited by high phytoplankton production that affected light environments[@greening2014; @johansson15].

@@

The second goal of the analysis was to evaluate if seagrass changes were linked to long-term changes in water temperature and salinity. The conceptual model for evaluating these changes describes the fundamental niche space where seagrass growth and reproduction is hypothesized to be greatest within optimal ranges for forcing factors that are present in the environment [@hutchinson1957; @vandermeer1972]. In the simplest form, this can conceptualized as a bell curve with optimal conditions defined within a range of values for a single parameter, where reduced growth or mortality is observed outside of these ranges. Because both water temperature and salinity were evaluated, the same model can be conceptualized in two-dimensional space ([@fig-concept]). Seagrass growth can be limited when temperature is below or above the optimum range, when salinity is below or above the optimum range, or when both temperature and salinity conditions are outside of the optimum range. Based on the results of the trend tests, we hypothesized that seagrasses are likely stressed by both high temperature and low salinity (bottom right, @fig-concept). Although the fundamental niche space can be defined in multiple dimensions for many parameters, we focus on water temperature and salinity given that other dominant forcing factors, i.e., light availability, have been relatively stable in recent years [@tbep0323].

A fundamental challenge describing niche space is identifying the boundaries for optimal conditions. In Tampa Bay, three dominant seagrass species occur: *Halodule wrightii* (shoal grass), *Syringodium filiforme* (manatee grass), and *Thalassia testudinum* (turtle grass) [@lewis1985; @phillips88]. These species co-occur often in mixed beds throughout the bay, although some differences in abundance are observed across salinity ranges. Shoal grass is tolerant of a wide range of salinity [@lirman2003], but is more abundant in oligo/mesohaline portions of Tampa Bay. Conversely, turtle grass is less tolerant of low salinity and is more abundant in more euryhaline conditions near the mouth of Tampa Bay. Reported salinity ranges for each of these species varies depending on location, season, and other co-occurring factors like temperature [@phillips1960; @mcmillan1967; @zieman1975; @lewis1985], although most studies place lower limits of salinity in the range of 15-25 ppt. Optimal temperature ranges are similar between these temperate-tropical species, with reduced growth observed at temperatures above 30$^\circ$ C [@zieman1975; @lewis1985].

Because of the uncertainty in defining *in situ* thresholds for optimal temperature and salinity ranges, multiple thresholds were evaluated to describe the potential for stress and how it may be related to changes in seagrass. Distinctions were not made between species, primarily due to lack of consensus between studies and likely site-specific ranges that affect seagrass growth in Tampa Bay, as well as challenges of modeling fundamental and realized niche spaces between competing species [@araújo2006]. First, we developed metrics of potential temperature and salinity stress by quantifying the maximum number of continuous days each year when temperature was above or salinity was below a given threshold. This approach assumed that stress could be observed based on duration of exposure (i.e., maximum number of continuous days each year) relative to a threshold that may or may not be outside of the optimum range for seagrasses. These metrics were quantified from the monthly long-term observations in the EPC data. To quantify daily counts each year, a continuous prediction of temperature and salinity over time at each of 45 stations was estimated using Generalized Additive Models (GAMs) fit to temperature or salinity with a single predictor for decimal year [@wood17]. Model fit for each station was considered sufficient to calculate daily predictions to assess potential stressor metrics (Figure S5, R$^2$ ranged from `r fitest$temp$minr2` to `r fitest$temp$maxr2` for temperature models, `r fitest$sal$minr2` to `r fitest$sal$maxr2` for salinity models, Tables S1, S2).

@@

For comparison to seagrass, the annual metrics calculated from the EPC data were referenced to approximate periods of time between the annual seagrass transect surveys, as opposed to the calendar year for describing trends above. Bay segment stressor metrics were calculated as the average counts in each "transect year" from all stations in each segment from 1998 to 2022. The stressor metrics were compared to frequency occurrence (all species) each year by bay segment. Generalized Linear regression Models (GLMs) were used to evaluate frequency occurrence in response to the independent variables, where the latter were the stressor metrics for temperature, salinity, both, bay segment, and time period (recovery pre 2016, decline post 2016). Two models were evaluated, one with the temperature and salinity metrics together and another with the both metric. Interactions were included between temperature, salinity, and time period or the both metric and time period, depending on the model. Bay segment was included as a main effect without interactions. The most parsimonious models were identified by forward and backward selection of the independent variables and comparison with AIC values [@sakamoto1986; @venables2002b]. All models excluded Lower Tampa Bay because of minimal seagrass change over time.

@@

The EPC dataset was also used to provide detailed information on station-level trends from 1975 to present (@fig-kendall, see Figure S4 for 1998 to 2022). All stations had significantly increasing temperature and decreasing salinity from 1975 to present, excluding one station in HB that did not have significantly decreasing bottom salinity (Figure [-@fig-kendall]a). Seasonally, most bay segments had more significant stations during the summer, early fall period for both temperature and salinity (Figure [-@fig-kendall]b, c), although some variation was observed throughout the bay. Temperature trends were more often observed in the summer or early fall for the upper bay segments (OTB, HB), whereas the lower bay segments (MTB, LTB) had more seasonally homogeneous trends. Seasonal trends in salinity did not vary as much between segments, although OTB and LTB had more stations with significantly decreasing trends than the other bay segments.

@@

Linear mixed-effects models for the EPC data showed similar trends in each bay segment for the number of days when temperature was above different thresholds. All of the temperature models for each of the three thresholds (29, 30, 31 $^\circ$C) showed increasing trends for each bay segment, with the largest slope of 1.4 days per year in OTB when temperature was above 29 $^\circ$C (Table S3). The estimated slopes for the number of days when temperature was above 30 $^\circ$C varied from 0.9 days per year in LTB to 1.1 days per year in MTB. Likewise, the mean number of days when temperature was above 30 $^\circ$C at the beginning and end of the period of record varied from 8 (OTB, HB) to 21 (LTB) days in 1975 to 55 (HB) to 63 (LTB) days in 2022 (Table S5, see Table S6 for 1998 to 2022). Trends for the other two temperature thresholds were similar.

The salinity models were less similar between bay segments compared to the temperature models, primarily because of the natural salinity gradient along the bay's longitudinal axis (Tables S3, S4). None of the bay segments had significantly increasing number of days per year when salinity was below 15 ppt. The only significant model for MTB was the number of days when salinity was below 25 ppt, which was increasing at a rate of 0.7 days per year. Both HB and OTB had significantly increasing number of days per year when salinity was below 20 and 25 ppt, although the slopes varied such that the rates in OTB were nearly double those in HB. The number of days per year when salinity was below 25 ppt in OTB increased by 2 days per year, whereas the number of days in HB increased by 0.9 days per year from 1975 to present. Likewise, the mean number of days per year when salinity was continuously below 25 ppt at the beginning and end of the period of record was 128 (OTB) and 131 (HB) days in 1975 and 171 (HB) and 219 (OTB) days in 2022 (Table S5, see Table S6 for 1998 to 2022).

@@

Global increases in temperature and altered precipitation patterns related to climate change have had measurable affects on the structure and functioning of a wide range of natural environments [@osland2015; @oliver2018]. For Tampa Bay, these changes have been demonstrated using long-term trends in water temperature and salinity, which mirrored long-term changes in air temperature and precipitation. Over the previous fifty years, Tampa Bay has gotten hotter and fresher; water temperature has increased by 0.03 - 0.07 $^\circ$C per year and salinity has decreased by 0.04 - 0.2 ppt per year. These changes were demonstrated in three long-term datasets with different sampling methods and periods of record. Understandably, the trends were most cearly observed in the dataset with the longest period of record (EPC), covering nearly fifty years of monthly observations, with the most dramatic changes occurring during the summer rainy season at locations farther from the mouth of the Bay. These long-term changes manifested into consistent trends in known seagrass stressors; the continuous number of days increased each year when temperature, salinity, or both crossed thresholds.
